**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-05-08 paper/manuscript.Rmd
ms updates

---
output:
  pdf_document:
    keep_tex: true
  word_document: default
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

**Other title options:**  
Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer
ASV level resolution is unnecessary for prediction of SRNs from microbiome data 
...

@@

## Introduction (~250 words)

Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States {}. Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e. colonoscopy), many people do not comply with recommended screening guidelines {} prompting a need for low cost, non-invasive detection methods. A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e. advanced adenomas) and cancer. Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on the amplicon sequencing of the 16S rRNA gene {}. However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown. Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs {}. However, whether the additional resolution provided by ASVs is useful for ML classification is unclear. Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. Topçuoğlu *et al* {mBio 2020} recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml). This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

## Results (~700 words)

@@

Utilizing publicly available 16S rRNA sequence data from patients with cancer and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quanitfied how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or cancer) using machine learning (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. [**TODO**: dig into literature on this: RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data]. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06). These results were consistent with ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels).  The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 2). **SUMMARY SENTANCE**

@@

Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods). As part of this step, features are removed that will not provide useful information to build the model. For example, strongly correlated features provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).

@@

## Discussion/Conclusions (~250 words)

@@

***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from NCBI SRA (accession #). This dataset contains stool samples from 490 subjects. Based on the available metadata, samples categorized as normal, high risk normal, or adenoma were labeled "normal" for this analysis and samples categorized as advanced adenoma or carcinoma were labeled as "screen relevant neoplasia" (SRN). This resulted in a total of 261 "normal" samples and 229 "SRN" samples.

@@

***Machine Learning.*** Machine learning models were run with the R package mikropml (v0.0.2) {} to predict the diagnosis category (normal vs SRN) of each sample. Data was preprocessed to normalize values (scale/center? range?), remove values with zero or near-zero variance, and collapse collinear features using default parameters. Initially the models were run with default hyperparameters, but were expanded if the peak performance was at the edge of the hyperparameter range. Each taxonomic model taxonomic level combination (e.g. Random Forest on genus) was run with 100 different seeds (1-100). Each seed split the data into a training (80%) and testing (20%) set, and output performance of the training and testing as area under the receiver operating curve (AUROC).

@@

![](../exploratory/figures/rf_with_stats_sideways.png){width=75%}

**Figure 1: Random Forest Model Performance with Significance.** Boxplots of AUROC values on the test dataset for 100 seeds predicting SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which is equivalent to random classification. Differences in AUROC values between taxonomic levels are significant unless otherwise specified with "NS" (not significant)

\newpage
## Tables

@@

![](../exploratory/figures/bin_prevalence.png){height=50%}  
**Supplemental Figure 2: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.

>>>>

---
bibliography: references.bib
link-citations: true
output:
  pdf_document:
    keep_tex: true
  word_document: default
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

**Other title options:**\
Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer\
...

@@

## Introduction (\~250 words)

Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States {}. Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e. colonoscopy), many people do not comply with recommended screening guidelines {} prompting a need for low cost, non-invasive detection methods. A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e. advanced adenomas) and carcinomas. Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}. However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown. Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs {}. However, whether the additional resolution provided by ASVs is useful for ML classification is unclear. Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml). This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

## Results (\~700 words)

@@

Utilizing publicly available 16S rRNA sequence data from patients with cancer and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quanitfied how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or cancer) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p \< 0.05) but not family level (p = 0.06). These results were consistent for ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels). The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 2). While the resolution provided by ASVs is useful in

@@

**is this paragraph useful???** *Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods). As part of this step, features are removed that will not provide useful information to build the model. For example, strongly correlated features provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).*

@@

## Discussion/Conclusions (\~250 words)

@@

***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from NCBI SRA (accession \#). This dataset contains stool samples from 490 subjects. Based on the available metadata, samples categorized as normal, high risk normal, or adenoma were labeled "normal" for this analysis and samples categorized as advanced adenoma or carcinoma were labeled as "screen relevant neoplasia" (SRN). This resulted in a total of 261 "normal" samples and 229 "SRN" samples.

@@

***Machine Learning.*** Machine learning models were run with the R package mikropml (v0.0.2) {} to predict the diagnosis category (normal vs SRN) of each sample. Data was preprocessed to normalize values (scale/center), remove values with zero or near-zero variance, and collapse collinear features using default parameters. Initially the models were run with default hyperparameters, but were expanded if the peak performance was at the edge of the hyperparameter range. Each taxonomic model taxonomic level combination (e.g. Random Forest on genus) was run with 100 different seeds (1-100). Each seed split the data into a training (80%) and testing (20%) set, and output performance of the training and testing as area under the receiver operating curve (AUROC).

@@

![](../exploratory/figures/rf_with_stats.png){width="75%"}

**Figure 1: Random Forest Model Performance with Significance.** Boxplots of AUROC values on the test dataset for 100 seeds predicting SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which is equivalent to random classification. Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.

## Tables

@@

![](../exploratory/figures/bin_prevalence.png){height="50%"}\
**Supplemental Figure 2: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.

## References

### 2021-05-08 paper/manuscript.tex
ms updates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\textbf{Other title options:}\\
Optimal Taxonomic Resolution for Microbiome-Based Classification of
Colorectal Cancer ASV level resolution is unnecessary for prediction of
SRNs from microbiome data \ldots{}

@@

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States \{\}.
Early detection and treatment are essential to increase survival rates,
but for a variety of reasons including the invasiveness and high cost of
screening (i.e.~colonoscopy), many people do not comply with recommended
screening guidelines \{\} prompting a need for low cost, non-invasive
detection methods. A growing body of research points to the gut
microbiome as a promising target for non-invasive screening to detect
screen relevant neoplasias (neoplasms?) (SRNs) consisting of
pre-cancerous polyps (i.e.~advanced adenomas) and cancer. Efforts to
realize the diagnostic potential of the gut microbiome in detecting SRNs
have focused on machine learning (ML) methods using abundances from
operational taxonomic unit (OTU) classifications based on the amplicon
sequencing of the 16S rRNA gene \{\}. However, whether this is the
optimal taxonomic resolution for classifying SRNs from microbiome data
is unknown. Additionally, recent work has pushed for the use of amplicon
sequence variants (ASVs) to replace OTUs for marker-gene analysis
because of the improved resolution with ASVs \{\}. However, whether the
additional resolution provided by ASVs is useful for ML classification
is unclear. Since ML classification relies on consistent differences
between groups, its possible that the resolution at the ASV level is too
individualized to accurately differentiate groups. Topçuoğlu \emph{et
al} \{mBio 2020\} recently demonstrated effective application of machine
learning (ML) to microbiome based classification problems and developed
a framework for applying ML practices in a more reproducible way
(mikropml). This analysis utilizes the reproducible framework developed
by Topçuoğlu \emph{et al} to quantify which ML method and taxonomic
level produce the best performing classifier for detecting SRNs from
microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from patients with
cancer and healthy controls, we generated abundance tables with Mothur
\{\} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV
levels. Using the taxonomic abundance data, we quanitfied how accurately
samples could be classified as ``normal'' or ``SRN'' (i.e.~advanced
adenoma or cancer) using machine learning (methods). Across the five
machine learning methods tested, model performance tended to increase
with taxonomic level usually peaking around genus/OTU level before
dropping off slightly with ASVs (Supplemental Figure 1). Random forest
(RF) was consistently the top performer at most taxonomic levels.
{[}\textbf{TODO}: dig into literature on this: RF might be more
appropriate for microbiome analysis since its more suitable for zero
inflated data{]}. Within the RF model, the highest AUROCs were observed
for family (median AUROC: 0.687), genus (median AUROC: 0.686), and OTU
(median AUROC: 0.698) level data with no significant difference between
the three (Figure 1). Performance with ASVs (median AUROC: 0.676) was
significantly lower than OTU and genus (p \textless{} 0.05) but not
family level P = 0.06). These results were consistent with ASVs
generated with DADA2 \{\} (median AUROC: 0.66). These results suggest
that finer taxonomic resolution is not necessarily better for
differentiating a persons cancer status from the microbiome.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus/OTU data there is enough data and variation but at the
ASV level, the data is too specific to individuals and doesn't overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa are present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa)
levels). The opposite is observed at the OTU and ASV level where 60\%
and 53\% of taxa respectively are only present in less than 25\% of
samples (Supplemental Figure 2). \textbf{SUMMARY SENTANCE}

@@

Of note, the ML pipeline includes a pre-processing step that occurs
prior to training and classifying the ML models (methods). As part of
this step, features are removed that will not provide useful information
to build the model. For example, strongly correlated features provide
the same information to build the model and thus can be collapsed.
Additionally, features with zero or near-zero variance will not help the
model differentiate groups and thus can be removed. Interestingly,
despite starting with 104106 features at the ASV level, only 478 (0.5\%)
remained after pre-processing. At the OTU level, 20079 of the 705
features (3.5\%) remained after preprocessing (Table 1).

@@

\textbf{\emph{Machine Learning.}} Machine learning models were run with
the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
(normal vs SRN) of each sample. Data was preprocessed to normalize
values (scale/center? range?), remove values with zero or near-zero
variance, and collapse collinear features using default parameters.
Initially the models were run with default hyperparameters, but were
expanded if the peak performance was at the edge of the hyperparameter
range. Each taxonomic model taxonomic level combination (e.g.~Random
Forest on genus) was run with 100 different seeds (1-100). Each seed
split the data into a training (80\%) and testing (20\%) set, and output
performance of the training and testing as area under the receiver
operating curve (AUROC).

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats_sideways.png}

\textbf{Figure 1: Random Forest Model Performance with Significance.}
Boxplots of AUROC values on the test dataset for 100 seeds predicting
SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which
is equivalent to random classification. Differences in AUROC values
between taxonomic levels are significant unless otherwise specified with
``NS'' (not significant)

@@

\begin{longtable}[]{@{}lrrc@{}}
\toprule
Taxonomic Level & Number of Features &
\makecell[c]{Number of Features \\ After Preprocessing} &
\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\ \addlinespace
\midrule
\endhead
Phylum & 19 & 9 & 47.4 \% \\ \addlinespace
Class & 36 & 19 & 52.8 \% \\ \addlinespace
Order & 65 & 28 & 43.1 \% \\ \addlinespace
Family & 124 & 54 & 43.5 \% \\ \addlinespace
Genus & 316 & 115 & 36.4 \% \\ \addlinespace
OTU & 20079 & 705 & 3.5 \% \\ \addlinespace
ASV & 104106 & 478 & 0.5 \% \\ \addlinespace
\bottomrule
\end{longtable}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\textbf{Other title options:}\\
Optimal Taxonomic Resolution for Microbiome-Based Classification of
Colorectal Cancer\\
\ldots{}

@@

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States \{\}.
Early detection and treatment are essential to increase survival rates,
but for a variety of reasons including the invasiveness and high cost of
screening (i.e.~colonoscopy), many people do not comply with recommended
screening guidelines \{\} prompting a need for low cost, non-invasive
detection methods. A growing body of research points to the gut
microbiome as a promising target for non-invasive screening to detect
screen relevant neoplasias (neoplasms?) (SRNs) consisting of
pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
realize the diagnostic potential of the gut microbiome in detecting SRNs
have focused on machine learning (ML) methods using abundances from
operational taxonomic unit (OTU) classifications based on amplicon
sequencing of the 16S rRNA gene \{\}. However, whether this is the
optimal taxonomic resolution for classifying SRNs from microbiome data
is unknown. Additionally, recent work has pushed for the use of amplicon
sequence variants (ASVs) to replace OTUs for marker-gene analysis
because of the improved resolution with ASVs \{\}. However, whether the
additional resolution provided by ASVs is useful for ML classification
is unclear. Since ML classification relies on consistent differences
between groups, its possible that the resolution at the ASV level is too
individualized to accurately differentiate groups. Topçuoğlu \emph{et
al} (\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
recently demonstrated effective application of machine learning (ML) to
microbiome based classification problems and developed a framework for
applying ML practices in a more reproducible way (mikropml). This
analysis utilizes the reproducible framework developed by Topçuoğlu
\emph{et al} to quantify which ML method and taxonomic level produce the
best performing classifier for detecting SRNs from microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from patients with
cancer and healthy controls, we generated abundance tables with Mothur
\{\} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV
levels. Using the taxonomic abundance data, we quanitfied how accurately
samples could be classified as ``normal'' or ``SRN'' (i.e.~advanced
adenoma or cancer) using five machine learning methods with the Mikropml
R package (methods). Across the five machine learning methods tested,
model performance tended to increase with taxonomic level usually
peaking around genus/OTU level before dropping off slightly with ASVs
(Supplemental Figure 1). Random forest (RF) was consistently the top
performer at most taxonomic levels. Within the RF model, the highest
AUROCs were observed for family (median AUROC: 0.687), genus (median
AUROC: 0.686), and OTU (median AUROC: 0.698) level data with no
significant difference between the three (Figure 1). Performance with
ASVs (median AUROC: 0.676) was significantly lower than OTU and genus (p
\textless{} 0.05) but not family level (p = 0.06). These results were
consistent for ASVs generated with DADA2 \{\} (median AUROC: 0.66).
These results suggest that finer taxonomic resolution is not necessarily
better for differentiating a persons cancer status from the microbiome.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus/OTU data there is enough data and variation but at the
ASV level, the data is too specific to individuals and doesn't overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa are present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa)
levels). The opposite is observed at the OTU and ASV level where 60\%
and 53\% of taxa respectively are only present in less than 25\% of
samples (Supplemental Figure 2). While the resolution provided by ASVs
is useful in

@@

\textbf{is this paragraph useful???} \emph{Of note, the ML pipeline
includes a pre-processing step that occurs prior to training and
classifying the ML models (methods). As part of this step, features are
removed that will not provide useful information to build the model. For
example, strongly correlated features provide the same information to
build the model and thus can be collapsed. Additionally, features with
zero or near-zero variance will not help the model differentiate groups
and thus can be removed. Interestingly, despite starting with 104106
features at the ASV level, only 478 (0.5\%) remained after
pre-processing. At the OTU level, 20079 of the 705 features (3.5\%)
remained after preprocessing (Table 1).}

@@

\textbf{\emph{Machine Learning.}} Machine learning models were run with
the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
(normal vs SRN) of each sample. Data was preprocessed to normalize
values (scale/center), remove values with zero or near-zero variance,
and collapse collinear features using default parameters. Initially the
models were run with default hyperparameters, but were expanded if the
peak performance was at the edge of the hyperparameter range. Each
taxonomic model taxonomic level combination (e.g.~Random Forest on
genus) was run with 100 different seeds (1-100). Each seed split the
data into a training (80\%) and testing (20\%) set, and output
performance of the training and testing as area under the receiver
operating curve (AUROC).

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats.png}

\textbf{Figure 1: Random Forest Model Performance with Significance.}
Boxplots of AUROC values on the test dataset for 100 seeds predicting
SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which
is equivalent to random classification. Significance between taxonomic
levels was quantified by comparing the difference in mean AUROC and is
denoted by letters A through E on the right side of the plot; taxonomic
levels with the same letter are in the same significance group and are
not significantly different from one another.

@@

\begin{longtable}[]{@{}lrrc@{}}
\toprule
Taxonomic Level & Number of Features &
\makecell[c]{Number of Features \\ After Preprocessing} &
\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\
\midrule
\endhead
Phylum & 19 & 9 & 47.4 \% \\
Class & 36 & 19 & 52.8 \% \\
Order & 65 & 28 & 43.1 \% \\
Family & 124 & 54 & 43.5 \% \\
Genus & 316 & 115 & 36.4 \% \\
OTU & 20079 & 705 & 3.5 \% \\
ASV & 104106 & 478 & 0.5 \% \\
\bottomrule
\end{longtable}

@@

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
Topçuoğlu, Begüm D., Nicholas A. Lesniak, Mack T. Ruffin, Jenna Wiens,
and Patrick D. Schloss. 2020. {``A Framework for Effective Application
of Machine Learning to Microbiome-Based Classification Problems.''}
\emph{mBio} 11 (3). \url{https://doi.org/10.1128/mBio.00434-20}.

\end{CSLReferences}
